In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-20 trunk/tex/sigmasfr/ms.tex
updates to section 3.3

We present the discovery of compact, obscured star formation in
galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  Using optical
morphologies from the Hubble Space Telescope and infrared photometry
from the Wide-field Infrared Survey Explorer, we estimate star
formation rate (SFR) surface densities that approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$.  This is comparable
to the Eddington limit from radiation pressure on dust grains.  We
argue that feedback associated with a compact starburst in the form of
radiation pressure from massive stars and ram pressure from supernovae
and stellar winds is sufficient to launch the high-velocity outflows
we observe, without the need to invoke feedback from an active
galactic nucleus.

@@

\subsection{The Outflow Launching Mechanism}

@@

The high-$\sigmasfr$ galaxies in our sample constitute a rare
population (see Figure~\ref{fig:sigmasfr}), suggesting that they
represent an unusual or short-lived phase.  Mergers of gas-rich
galaxies are a viable mechanism for producing compact starbursts
(e.g., Mihos+94), and such gas-rich major mergers are rare at
$z\sim0.6$ due to the decline in both the merger rate and the gas
fraction of galaxies since $z\sim1$.  The high-$\sigmasfr$ galaxies in
particular are in a phase where there is strong feedback but star
formation has not been quenched.  The length of this phase may be set
by the gas consumption timescale or the timescale for feedback to
suppress subsequent star formation.  Based on the Kennicutt--Schmidt
(K-S) relation \citep{ken98}, a galaxy with $r_e\approx100$~pc,
$\textnormal{SFR}\approx300$~\msun, and
$\sigmasfr=5000$~\units\ (assuming half of the star formation occurs
within the effective radius) would have a gas surface density of
$\Sigma_{gas}\sim2\times10^5$~\msun~pc$^{-2}$, corresponding to
$M_{gas}\sim5\times10^{9}$~\msun\ inside 100~pc, which would be
consumed on a timescale $\sim30$~Myr.  It would be worthwhile to
directly measure the molecular gas masses and kinematics of these
extreme sources with CO observations to test this scenario.

The vertical sequence in Figure~\ref{fig:sigmasfr} may be related to
an evolutionary sequence, with the smaller $\sigmasfr$ galaxies being
further along in their evolution towards a post-starburst phase.
However, it is interesting to note that all 12/29 galaxies above the
$\sigmasfr\approx45$~\units\ \citet{meu97} limit exhibit outflows
(with median centroid velocity $v=-1500$~\kms), while all 7/29
galaxies without detected outflows have $\sigmasfr<30$~\units.  If a
compact starburst is a requirement for the production of high-velocity
outflows, it may be that these latter sources have not gone through
such a phase (see A. Robaina, et al, in preparation for a discussion
of the relationship between outflow velocity, galaxy morphology, and
stellar population age).

@@

It is also worhwhile to consider the implications of these galaxies
for models of massive galaxy formation.  Models do not produce
starbursts this small.  Models that produce ellipticals have to invoke
AGN feedback.  Models for size evolution would allow these galaxies to
reach the size--mass relation by $z=0$.

@@

@@

Sample as a whole. Interesting that high-$\sigmasfr$ galaxies tend to
have larger outflow velocities.

@@

>>>>

We present the discovery of compact, obscured star formation in
galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  Using optical
morphologies from the Hubble Space Telescope and infrared photometry
from the Wide-field Infrared Survey Explorer, we estimate star
formation rate (SFR) surface densities that approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$, which is comparable
to the Eddington limit from radiation pressure on dust grains.  We
argue that feedback associated with a compact starburst in the form of
radiation pressure from massive stars and ram pressure from supernovae
and stellar winds is sufficient to produce the high-velocity outflows
we observe, without the need to invoke feedback from an active
galactic nucleus.

@@

\subsection{The Outflow Launching Mechanism}\label{sec:launch}

@@

The high-$\sigmasfr$ galaxies in our sample constitute a rare
population (see Figure~\ref{fig:sigmasfr}), suggesting that they
represent an unusual or short-lived phase.  Mergers of gas-rich
galaxies are a viable mechanism for producing compact starbursts
(e.g., Mihos+94), but such gas-rich major mergers are rare at
$z\sim0.6$ due to the decline in both the merger rate and the gas
fraction of galaxies since $z\sim1$.  The high-$\sigmasfr$ galaxies
are also in a particular phase where there is strong feedback but star
formation has not been quenched.  The length of this phase may be set
by the gas consumption timescale or the timescale for feedback to
suppress subsequent star formation.  Based on the Kennicutt--Schmidt
(K-S) relation \citep{ken98}, a galaxy with $r_e\approx100$~pc,
$\textnormal{SFR}\approx300$~\msun, and
$\sigmasfr=5000$~\units\ (assuming half of the star formation occurs
within the effective radius) would have a gas surface density of
$\Sigma_{gas}\sim2\times10^5$~\msun~pc$^{-2}$, corresponding to
$M_{gas}\sim5\times10^{9}$~\msun\ inside 100~pc, which would be
consumed on a timescale $\sim30$~Myr.  Such estimates could be tested
with observations of molecular gas masses and kinematics for these
extreme galaxies.

% which rely on extrapolating the K-S relation beyond the range where it had been calibrated,

The vertical sequence among the galaxies in our sample in
Figure~\ref{fig:sigmasfr} could be related to an evolutionary
sequence, with smaller $\sigmasfr$ galaxies being further along in
their evolution towards a post-starburst phase.  However, it is
interesting to note that all 12/29 galaxies above the \citet{meu97}
limit ($\sigmasfr\approx45$~\units) exhibit outflows (with median
centroid velocity $v=-1500$~\kms), while all 7/29 galaxies without
detected outflows have $\sigmasfr<30$~\units.  If a compact starburst
is a requirement for the production of high-velocity outflows, as
suggested by previous the discussion in Section~\ref{sec:launch}, it
may be that the non-outflow sources have not gone through such a phase
(see A. Robaina, et al., in preparation for a discussion of the
relationship between outflow velocity, galaxy morphology, and stellar
population age).

@@

Finally, it is worthwhile to consider the implications of these
galaxies for models of massive galaxy formation.  While simulaitons of
major galaxy mergers with large gas fractions ($f_{gas}\sim50$\%) can
produce $\textnormal{\mstar}\sim10^{11}$~\msun\ remnants with
effective radii $r_{e}\sim1$~kpc \citep[e.g.,][]{wuy10}, similar to
the observed sizes of compact quiescent galaxies at $z\sim2$
\citep[e.g.,][]{van08}, our sample includes galaxies of similar mass
that are smaller by almost an order of magnitude.  If these systems do
evolve into quiescent elliptical galaxies, it would be quite
challenging for them to grow in size from $r_e\sim0.1$~kpc to
$r_e\sim5$~kpc to reach the local size--mass relation in the
$t\sim6$~Gyr since $z=0.6$.  In this context, it is relevant point to
note that the half-light radius (ignoring dust attenuation) at
rest-frame $V$ band (where our HST observations prbe) can be a factor
of 5--10 smaller than the half-mass radius for a system near the peak
of the starburst \citep{wuy10}.  It would be worthwhile to test for
such size discrepancies and to pronbe the radial dependence of the
mass-to-light ratio by measuring sizes at rest-frame near-IR
wavelengths.

@@

%Sample as a whole. Interesting that high-$\sigmasfr$ galaxies tend to
%have larger outflow velocities.

Their descendants could perhaps be the local galaxies that are
 significant outliers from the size--mass relation.

Models that produce ellipticals have to invoke AGN feedback to quench
star formation.

@@

@@

\bibitem[Wuyts et al.(2010)]{wuy10} Wuyts, S., Cox, T.~J., Hayward,
  C.~C., et al.\ 2010, \apj, 722, 1666
